---
title: "2/12/19 notes"
author: "pwilliams"
date: "2/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Using tidyverse, we receive access to a bunch of new commands
```{r include=FALSE}
library(tidyverse)
read.delim("./Exams/Exam_1/DNA_Conc_by_Extraction_Date.csv", sep = "\t") -> df
```


#02/12/19 Notes



## Piping commands using tidyverse packages
 **%>%** works similar as a pipe command. Select lets us choose columns (a - in front of column means everything but that column)

```{r}
# First we call our data frame, then we pipe our dataframe into the next commmand, in this case, the select() command.
selectYC = df %>%
  select(Year_Collected)
head(selectYC)
```

 filter() lets you access all the rows of a specifc command (in this case all the rows with year collected > 2005)
```{r}
filter1 = df %>%
  filter(Year_Collected > 2005) %>%
  mutate(k2 = DNA_Concentration_Katy^2)
head(filter1)
  
```

The group_by() function allows us to create groups within columns. When paired with summarise it becomes **DEADLY**
```{r}
gbsum1 = df %>%
  group_by(Year_Collected) %>%
  summarise(N = n(), MeanBen = mean(DNA_Concentration_Ben), SDBen = sd(DNA_Concentration_Ben))
head(gbsum1)
#This grouped our Year_Collected column nicely
```

```{r include=FALSE}
data("iris")
df = iris
```
We practiced subsetting large data frames using tidyr commands with the Iris data set


```{r}
df2 = df %>%
  group_by(Species) %>%
  summarise(N = n(), SLMEAN = mean(Sepal.Length), SDsl = sd(Sepal.Length), MINsl = min(Sepal.Length), MAXsl = max(Sepal.Length)) %>%
  as.data.frame() # The as.data.frame() function needed to be added because, without it, the dataframe is stored as a tibble, which makes it difficult to work with in some programs or on some devices

#Check our new class of df2
class(df2)
df2
```

Next we used how to search within dataframes (specifically column and row names) using the select() function and its arguments.
```{r}

#Starts_with selects column names that start with a provided search factor
startwithS = df %>% 
  select(starts_with("S"))
head(startwithS)


#To explore data sets when column names are unknown:
pickone = df%>%
  select(one_of(c("Petal.Length", "Sepal.Roundness")))
head(pickone)
```
Here are some common commands for tidying and manipulating data in tidyverse:
+arrange() 
+filter()
+select()
+mutate()
+group_by()



##Tidying data sets using tidyr

In tidy data:
 + Every row = observation
 + Every column = single variable
 + Element = Value
 
 
Here are two data frames
```{r echo=FALSE}
df1 = data.frame(CustomerId = c(1:6), Product = c(rep("Toaster", 3), rep("Radio", 3)))
df2 = data.frame(CustomerId = c(2, 4, 6), State = c(rep("Alabama", 2), rep("Ohio", 1)))

df1
df2
```

Join them together using the commands below:
```{r}
full_join(df1,df2, by="CustomerId")
#Combines everything from both dfs
left_join(df1,df2, by="CustomerId")
#combines everything from the by argument in left df with matching by arguments in right df
right_join(df1,df2, by="CustomerId")
##combines everything from the by argument in left df with matching by arguments in right df
inner_join(df1,df2, by="CustomerId")
semi_join(df1,df2, by="CustomerId")
anti_join(df1,df2, by="CustomerId")
```
What is the difference between these different functions?

##Converting wide data sets into long ones using the gather() function. 
```{r include=FALSE}
read.csv("..//Data_Course/Data/Utah_Religions_by_County.csv") -> utah
```

```{r}
head(utah)
```
This is an example of a wide data set. Each variable that consists of a relgion should be condensed into two columns: one titled Religion, and the other titled Proportion. This can be done using the gather() command:
```{r}
utah.long = gather(utah, key = "Religion", value = "Proportion", -c(1:4))

utah.long$County <- factor(str_remove(utah.long$County, " County"))
# We simply removed the word County from every element within the County variable
```

Now we can plot this data set using ggplot (which we could not do before)
```{r}
ggplot(utah.long, aes(x = Religion, y = Proportion, fill = Religion)) +
  geom_boxplot() + coord_flip() +facet_wrap(utah.long$County)
```
Hmmm, it looks like the observation "LDS" is having a large affect on the rest of our data. Lets get rid of that variable just to see what the other data looks like.
```{r}
utah.long %>%
  filter(Religion != "LDS") %>%
  ggplot(aes(x = Religion, y = Proportion, fill = Religion)) +
  geom_boxplot() + coord_flip() + facet_wrap(~County)
```