---
title: "02/19/19 notes"
author: "pwilliams"
date: "2/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 02/19/19 Notes  
## Working with Excel/ Tidying data  

```{r, include=FALSE}
library(tidyverse)
# First and foremost, download needed packages. For today, tidyverse was sufficient ####
```

### Ugly Data Excersize 1

The first thing we did was upload a real life example of ugly data and made it a long data set:  
```{r message=FALSE, warning=FALSE}
df = read.csv("./Exercises/ugly_data1.csv")
long = gather(df, Group, Value, 2:9)
```

Now that our data was in the right format, we can start to tidy it up a bit! The first step is to fix that horrible *Groups* column. It looks like we need to split the gender from the age-range, but there is nothing separatin them so using our separate() function might prove a little difficult as is. There are two ways to fix this:  
The first, looks like this:
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
malerows = grep("^m", long$Group) # This will find any string in the vector that starts with m
femalerows = grep("^f", long$Group) # This will find any string in the vector that starts with f
long$Gender[malerows] <- "Male" # This will add the value "Male" in a newly created column "Gender" for all values that start with m in the group column
long$Gender[femalerows] <- "Female" #This will do the same but for Female
long$Group = str_remove(long$Group, "^m")  # This will get rid of the m at the beginning of the string
long$Group = str_remove(long$Group, "^f")  # This will get rid of the f at the beginning of the string
long$Group = str_replace(long$Group, "014", "0014") # Replacing our 014 with 0014 allows us to neatly insert a dash after the second character of our string 


######## This is where we insert in the dash ########
ages1 = substr(long$Group, start = 1, stop = 2)
ages2 = substr(long$Group, start = 2, stop = 4)
ages = paste0(ages1, "-", ages2)
```

The second way, a little less coding involved, looks like this: 
```{r}
group1 = substr(long$Group, start = 1, stop = 1)
group2 = substr(long$Group, start = 2, stop = 5)
groups = paste0(group1, "-", group2)
long$Group = groups
long = separate(long, Group, into = c("Gender", "AgeRange"), sep = "-")
```






